\documentclass{ximera}

\title{Multivariable Optimization: First Derivatives}
\author{Zack Reed}

\begin{document}
\begin{abstract}
In this activity we extend single-variable optimization techniques to multivariable settings.
\end{abstract}
\maketitle


\subsection*{Generalizing to Higher Dimensions: From First Derivatives to The Gradient Vector}

\begin{definition}
For a function $f(x,y)$, the \textbf{gradient} is the vector of partial derivatives:
$$\nabla f = \left\langle \frac{\partial f}{\partial x}, \frac{\partial f}{\partial y} \right\rangle$$
\end{definition}

In single-variable calculus, we found points where $f'(x) = 0$ to locate possible min and max values. This was because nonzero derivatives meant the function \wordChoice{\choice{possibly increased or decreased}\choice[correct]{had to either increase or decrease}\choice{was constant}} at that point.

Similarly, in multivariable calculus, if the gradient $\nabla f$ is nonzero at a point, then the function \wordChoice{\choice{possibly increases or decreases}\choice[correct]{must either increase or decrease}\choice{is constant}} in some direction at that point. This is because the directional derivative \wordChoice{\choice{is independent of the gradient}\choice[correct]{depends on a dot product with the gradient}}.

So, the only possible locations for maximum or minimum values are \wordChoice{\choice{where the gradient is nonzero}\choice[correct]{where the gradient is zero}}.

\begin{definition}
A point $(x_0, y_0)$ is a \textbf{critical point} of $f(x,y)$ if:
$$\nabla f(x_0, y_0) = \langle 0, 0 \rangle$$.

This means:
$$\frac{\partial f}{\partial x}(x_0, y_0) = 0 \quad \text{and} \quad \frac{\partial f}{\partial y}(x_0, y_0) = 0$$
\end{definition}

\begin{remark}
Technically, we also care about when the derivative (or graident) is undefined, but for this activity we will focus on where it is zero.
\end{remark}

\begin{problem}
Why do we care about critical points?

\begin{multipleChoice}
    \choice{They are always minima}
    \choice{They are always maxima}
    \choice[correct]{Local extrema must occur at critical points (if they exist in the interior)}
    \choice{The function is undefined at critical points}
\end{multipleChoice}
\end{problem}

\begin{problem}
Find the critical points of $f(x,y) = x^2 + y^2 - 4x + 6y + 5$.

First, compute the partial derivatives:
$$\frac{\partial f}{\partial x} = \answer{2x - 4}$$
$$\frac{\partial f}{\partial y} = \answer{2y + 6}$$

Set them equal to zero and solve:
$$2x - 4 = 0 \Rightarrow x = \answer{2}$$
$$2y + 6 = 0 \Rightarrow y = \answer{-3}$$

The critical point is $(\answer{2}, \answer{-3})$.

This is a paraboloid (bowl shape). Based on the shape, we can guess this critical point is a \wordChoice{\choice{maximum}\choice[correct]{minimum}\choice{saddle point}}.

%plot the function
\begin{center}
\begin{tikzpicture}
    \begin{axis}[
        axis lines = center,
        xlabel = $x$,
        ylabel = {$y$},
        title = {Graph of $f(x,y) = x^2 + y^2 - 4x + 6y + 5$},
        width=10cm,
        height=8cm,
        grid=major,
        view={60}{30},
    ]
    \addplot3[surf,domain=-1:5,y domain=-5:1,samples=20] {x ^2 + y ^2 - 4*x + 6*y + 5};
    \end{axis}
\end{tikzpicture}
\end{center}

\end{problem}

\section*{Caution: Not All Critical Points are Extrema!}

In single-variable calculus, we weren't always guaranteed a max or min at critical points. The same is true in multivariable calculus. Let's explore a common phenomenon where a critical point exists but no maximum or minimum occurs, called a \emph{saddle point}.


\begin{problem}
Find the critical points of $f(x,y) = x^2 - y^2$.

The partial derivatives are:
$$\frac{\partial f}{\partial x} = \answer{2x}$$
$$\frac{\partial f}{\partial y} = \answer{-2y}$$

Setting both equal to zero: $x = \answer{0}$ and $y = \answer{0}$.

The critical point is $(\answer{0}, \answer{0})$.

This function looks like a saddle at the origin - it curves up in one direction and down in another. The critical point is neither a minimum nor a maximum! This is called a saddle point.

\end{problem}

\begin{center}
\begin{tikzpicture}
    \begin{axis}[
        axis lines = center,
        xlabel = $x$,
        ylabel = {$y$},
        title = {Graph of $f(x,y) = x^2 - y^2$},
        width=10cm,
        height=8cm,
        grid=major,
        view={60}{30},
    ]
    \addplot3[surf,domain=-2:2,y domain=-2:2,samples=20] {x ^2 - y ^2};
    \end{axis}
\end{tikzpicture}
\end{center}


\end{document}
